Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-06-18 paper/paperv3.qmd
saving output of breeding success fig

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10,
         recruit_age = ifelse(max(year) < 2020 && any(observed_int == 1),
                              min(age[observed_int == 1]),
                              NA)) %>% 
  ungroup()

@@

@@

## Selective appearance and disappearance

```{r}
#| label: selective_values

recruit_by_6 <- mean(sealdat$recruit_age <= 6, na.rm = TRUE)

Age-dependent life history shifts can result from within- or between-animal changes [@vandepol2006]. We hypothesized age-dependent changes in reproduction were due to senescence, i.e. within-animal changes. However, it is possible that selective appearance or disappearance of animals in the population can create the same pattern. For example, if animals the breed more frequently have lower survival, then older animals will have lower reproductive traits not because of senescence but rather because only infrequent breeders survive to older ages (i.e. selective disappearance). We tested whether our results were influenced by selective disappearance, following the methods @vandepol2006, by adding a term for longevity to our models. We did not test for selective appearance because delayed sexual maturity was very rare in this population. `r scales::label_percent(accuracy = 0.1)(recruit_by_6)` of seals recruited at ages 4, 5, or 6.

```{r}
#| label: sel-dis-h1

breed_sel_dis <- glmer(
  observed_int ~ age10 * age_cat + longevity + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

summary(breed_sel_dis)$coefficients

longevity_est <- summary(breed_sel_dis)$coefficients["longevity", "Estimate"]
longevity_se <- summary(breed_sel_dis)$coefficients["longevity", "Std. Error"]
longevity_coef <- sprintf(
  "%0.5f (95%% CI: %0.5f - %0.5f)", 
  longevity_est,
  longevity_est - 1.96 * longevity_se,
  longevity_est + 1.96 * longevity_se
)

longevity_p <- sprintf(
  "%0.3f", 
  summary(breed_sel_dis)$coefficients["longevity", "Pr(>|z|)"]
)

Selective disappearance did not explain the observed pattern of declining breeding probabilities with age. The coefficient for longevity was estimated at `r longevity_coef`, which was not significantly different than 0 ($p$ = `r longevity_p`). The predicted breeding probabilities for seals with longevity of 5, 10, and 17 years are shown in @fig-sel-dis-h1.

```{r}
#| label: fig-sel-dis-h1
#| fig-cap: Longevity did not explain changes in breeding probabilities with age. Predicted breeding probabilities for seals with longevity of 5, 10, and 17 years shown with solid, dotted, and dashed lines, respectively. 95% confidence intervals shown in shaded regions.

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(
  breed_sel_dis, 
  terms = c("age10 [all]", "age_cat", "longevity [5,10,17]")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old")),
         longevity = facet,
         longevity_fct = factor(longevity)) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

ggplot(breed_pred_pop, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, 
                  group = interaction(age_cat, longevity_fct),
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.25) +
  geom_line(aes(color = age_cat, 
                linetype = longevity_fct)) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

>>>>

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10) %>% 
  ungroup()

@@

ggsave("outputs/breeding_success.svg", height = 6, width = 10, units = "in")
```

@@
